---
title: "Projecte AD"
output: html_document
date: '2023-03-18'
---

```{r, warning=FALSE}
library(ggplot2)
library(ggfortify)
library(forecast)
library(plotly)
library(TSstudio)
library(zoo)
library(xts)
library(lubridate)
library(magrittr)
```

## 1. Model identification

```{r}
setwd(getwd())
ser=ts(read.table("export.dat")/1000,start=1999,freq=12)
```

```{r}
#autoplot(ser) + ggtitle("Total exportations in Spain") + labs(x = "Time", y = "Milions of euros")
ts_decompose(ser) 
```
```{r}
autoplot(ser) +
  labs(title = "Total Exportations in Spain", 
       x = "Time", 
       y = "Thousands Million Euros") +
  theme(plot.title = element_text(color = "black", size = 18, face = "bold", hjust=0.5),
        axis.title.x = element_text(color = "black", size = 14),
        axis.title.y = element_text(color = "black", size = 14))

```

```{r}
ts_seasonal(trser, type = "normal", title = "Seasonality")
```

```{r}
library(forecast)
forecast::Acf(d1d12trser)
```


```{r}
ts_cor(ser, lag.max = 72)
```

```{r}
ts.plot_ly(ser)
```


```{r}
check_res(m1)
```


```{r}
boxplot(ser~floor(time(ser)))
```

```{r}
# create mean and variance vectors
m <- apply(matrix(ser, nrow = 12, byrow = FALSE), 2, mean)
v <- apply(matrix(ser, nrow = 12, byrow = FALSE), 2, var)

# create data frame
df <- data.frame(m = m, v = v)

# create scatter plot with color mapped to variance
ggplot(df, aes(x = m, y = v)) +
  geom_point(color = "#2c3e50") +  # puntos del gráfico
  labs(title = "Relación entre la media y la varianza",
       x = "Mean",
       y = "Variance") +  # etiquetas de los ejes y título
  theme_classic()  # estilo del gráfico
```


```{r}
lambda = BoxCox.lambda(ser, lower=-1, upper=2)
trser = (ser^lambda - 1)/lambda
  boxplot(trser~floor(time(trser)))
# Discrete colors
a + scale_fill_brewer(palette="Dark2") + theme_minimal()
# create mean and variance vectors
m <- apply(matrix(trser, nrow = 12, byrow = FALSE), 2, mean)
v <- apply(matrix(trser, nrow = 12, byrow = FALSE), 2, var)

# create data frame
df <- data.frame(m = m, v = v)

# create scatter plot with color mapped to variance
ggplot(df, aes(x = m, y = v)) +
  geom_point(color = "#2c3e50") +  # puntos del gráfico
  labs(
       x = "Mean",
       y = "Variance") +  # etiquetas de los ejes y título
  theme_classic()  # estilo del gráfico
```

```{r}
plot(trser)
```

```{r}
ggmonthplot(trser)
```

```{r}
d12trser = diff(trser, 12)
#ggmonthplot(d12trser)
ts_seasonal(d12trser, type = "normal", title = "Seasonality")
```

```{r}
ts_plot(d12trser,
        title = "Total exportations in Spain transformed series",
        Xtitle = "Time",
        width = 3) %>%
  layout(
        xaxis = list(tickfont = list(size = 10)), 
        yaxis = list(tickfont = list(size = 10)))
         
```

```{r}
d1d12trser = diff(d12trser)
d1d1d12trser = diff(d1d12trser)
ts_plot(d1d12trser,
        title = "Total exportations in Spain transformed series",
        Xtitle = "Time",
        width = 3) %>%
  layout(
        xaxis = list(tickfont = list(size = 10)), 
        yaxis = list(tickfont = list(size = 10)))
```

```{r}
var(d12trser)
var(d1d12trser)
var(d1d1d12trser)
```

### Model proposal

```{r}
par(mfrow=c(1,2))
acf(d1d12trser, yli=c(-1,1), lag.max=72, lwd=2, col=c(2, rep(1,11)))
pacf(d1d12trser, yli=c(-1,1), lag.max=72, lwd=2, col=c(rep(1,11), 2))
par(mfrow=c(1,1))
```

```{r}
ts_cor(d1d12trser, lag.max = 72)

```



PARTE ESTACIONAL:
MA(6) MA(2) AR(2)

PARTE REGULAR:
AR(2)

m1: ARIMA(2,1,0)(0,1,6)
m2: ARIMA(2,1,0)(0,1,2)
m3: ARIMA(2,1,0)(2,1,0)

#### auto.arima

```{r}
print_t_ratio <- function(m1) {
  t_ratios <- round(m1$coef / sqrt(diag(m1$var.coef)), 2)
  cat("\nT-ratios:", t_ratios)
  
  significant <- abs(t_ratios) > 2
  cat("\nSignificant?:", significant)
  
  # Find parameter with lowest absolute T-ratio
  abs_t_ratios <- abs(t_ratios)
  if (any(abs_t_ratios < 2)) {
    min_index <- which.min(abs_t_ratios)
    if (abs_t_ratios[min_index] < 2) {
      cat("\nParameter with lowest T-ratio:", t_ratios[which.min(abs_t_ratios)], 
          " (index", min_index, ")")
    }
  } else {
      cat("\nAll the parameters are significant")
    }
}
```

```{r}
am1 = auto.arima(trser, max.p = 7, max.q = 7, max.P = 7, max.Q=7, ic="aic", d = 1, D = 1)
am2 = auto.arima(trser, max.p = 7, max.q = 7, max.P = 7, max.Q=7, ic="aicc", d = 1, D = 1)
am3 = auto.arima(trser, max.p = 7, max.q = 7, max.P = 7, max.Q=7, ic="bic", d = 1, D = 1)
```
```{r}
am1
am2
am3
```



#### m1

```{r}
(m1 = arima(d1d12trser,order=c(2,0,0),
           seasonal=list(order=c(0,0,6), period=12)))
print_t_ratio(m1)
```
```{r}
(m1 = arima(trser,order=c(2,1,0),
           seasonal=list(order=c(0,1,6), period=12)))
print_t_ratio(m1)
```
```{r, warning=FALSE}
(m1 = arima(trser,order=c(2,1,0),
           seasonal=list(order=c(0,1,6), period=12),
           fixed=c(NA,NA,NA,NA,NA,NA,0,NA)))
print_t_ratio(m1)
```
AIC improves => we keep this model

#### m2

```{r, warning=FALSE}
(m2 = arima(d1d12trser,order=c(2,0,0),
           seasonal=list(order=c(0,0,2), period=12)))
print_t_ratio(m2)
```
```{r, warning=FALSE}
(m2 = arima(trser,order=c(2,1,0),
           seasonal=list(order=c(0,1,2), period=12)))
print_t_ratio(m2)
```
All parameters are significant => we keep this model

#### m3

```{r}
(m3 = arima(d1d12trser,order=c(2,0,0),
           seasonal=list(order=c(2,0,0), period=12)))
print_t_ratio(m3)
```
```{r}
(m3 = arima(trser,order=c(2,1,0),
           seasonal=list(order=c(2,1,0), period=12)))
print_t_ratio(m3)
```
bad AIC compared with m1 and m2, thus we dont consider this model.

## 2. Residual analisis
### 1) Variance analisis

```{r}
resi = resid(m2);
plot(resi);
abline(h=0);
abline(h=c(-3*sd(resi),3*sd(resi)),lty=3,col=4)
```


```{r}
scatter.smooth(sqrt(abs(resi)), lpars=list(col=2))
```

### 2) Normality

```{r}
qqnorm(resi); qqline(resi,col=2,lwd=2)
```


```{r}
hist(resi,breaks=20, freq=FALSE)
curve(dnorm(x, mean=mean(resi),sd=sd(resi)), col=2, add=T)
```

```{r}
shapiro.test(resi)
```

### 3) Residual independence
```{r}
par(mfrow=c(1,2))
acf(resi,ylim=c(-1,1),lag.max=72,lwd=2,col=c(2,rep(1,11)))
pacf(resi,ylim=c(-1,1),lag.max=72,lwd=2,col=c(rep(1,11),2))
par(mfrow=c(1,1))
```

```{r}
tsdiag(m1,gof.lag=72)
```
### 4) AR and MA polynomials

```{r}
Mod(polyroot(c(1,-m1$model$phi)))
min(Mod(polyroot(c(1,-m1$model$phi))))
```

Totes les arrels del polinomi característic de la part AR tenen modulo superior a 1, per la qual cosa el model és causal

```{r}
Mod(polyroot(c(1,m1$model$theta)))
min(Mod(polyroot(c(1,m1$model$theta))))
```
otes les arrels del polinomi característic de la part MA. tenen modulo superior a 1, per la qual cosa el model és invertible

```{r,warning=FALSE}
plot(m1)
```



